Bioverse: The Habitable Zone Inner Edge Discontinuity as an Imprint of Runaway Greenhouse Climates on Exoplanet Demographics

### 2023-05-01 src/tex/ms.tex
plot radius evolution via ms workflow

% Author list
\author[0000-0001-8355-2107]{Martin Schlecker}
\affiliation{Department of Astronomy/Steward Observatory, The University of Arizona, 933 North Cherry Avenue, Tucson, AZ 85721, USA}
%\author[0000-0002-3286-7683]{Tim Lichtenberg}
%\affiliation{Kapteyn Astronomical Institute, University of Groningen, PO Box 800, 9700 AV Groningen, The Netherlands}
\author{al.}

% Abstract
\begin{abstract}
Long-term magma ocean phases on rocky exoplanets orbiting inside of the runaway greenhouse threshold -- the inner edge of the classical habitable zone -- may offer insights into the physical and chemical processes that distinguish potentially habitable worlds from others.
Thermal stratification of super-runaway planets is expected to significantly inflate their radii, potentially providing observational access to the runaway greenhouse transition in the form of a ``habitable zone inner edge discontinuity'' in radius--density space.
Here, we assess the ability of ground- and space-based telescopes to test this hypothesis using \bioverse, a simulation framework that leverages the combination of contextual information from the overall planet population with a survey simulator and a hypothesis testing framework.

We find that the demographic imprint of the runaway greenhouse transition is likely detectable with high-precision transit photometry for planet samples of size $\gtrsim$~100 if at least \SI{\sim 10}{\percent} of planets inside of the habitable zone inner edge harbor super-runaway climates.
Our survey simulations suggest that in the near future, ESA's \plato\ mission will be the most promising survey to probe the habitable zone inner edge discontinuity.
We determine survey strategies that maximize the statistical power of the obtained data, and we identify as key mission design drivers: 1. A follow-up campaign of planetary mass measurements and 2. The fraction of low-mass stars in the target sample.
Observational constraints on the runaway climate transition will provide crucial insights into the distribution of atmospheric volatiles among rocky exoplanets, which may help to identify the nearest potentially habitable worlds.
\end{abstract}

% Main body
\section{Introduction}
%\todo[inline]{papers: \citep[][introduce runaway GH]{} \citep[][statistical detection of GH transition]{Turbet2019}, \citep[][(initial) water content of terrestrial planets]{Lichtenberg2019} \citep[][water loss, O2 buildup around M dwarfs]{Luger2015}, \citep[][Solubility of H in magma oceans]{Hirschmann2012,SchlichtingShahar_inreview,} \citet{Hamano2013,Hamano2015,Barth2021}.}

Despite recent advancements in observational techniques, our understanding of terrestrial-sized planets remains woefully limited, with fundamental aspects of their nature, composition, and potential habitability still largely unknown.
Due to the inherent detection biases of current exoplanet detection techniques, the best-studied category of rocky exoplanets at present is that of hot or warm, close-in planets~\citep{2019AREPS..47..141J,2021JGRE..12606639B}.
These bodies experience conditions that are in some aspects comparable to the environment of the inner Solar System bodies at early stages of their evolution~\citep{Ikoma2018,2021ChEG...81l5735C}, which likely profoundly affected the distribution of volatiles between planetary core, mantle, and atmosphere.
Studies of the geophysical state of hot exoplanets can thus inform our understanding of the early evolutionary stages of Earth and other habitable worlds~\citep{Lichtenberg2022,Krijt2022}.

An example from the Solar System for the potential significance of these early stages are the divergent atmospheric evolutions of Venus and Earth~\citep[e.g.,][]{2019JGRE..124.2015K,2021JGRE..12606643K}.
While having accreted from a similar mass reservoir~\citep{2020plas.book..287R,2020SSRv..216...55K,2020SSRv..216...27M,2020plas.book....3Z} and despite their similar bulk properties~\citep{Smrekar2018}, they evolved into planets with very different surface conditions~\citep{1982Sci...216..630D,Kasting1988,Hamano2013,Kane2014,Way2020,Turbet2021}.
Both planets underwent a giant impact phase~\citep{2020plas.book..287R,2020NatGe..13..265G,Liu2022} that melted their mantles~\citep{2012AREPS..40..113E,2018RSPTA.37680109S,Lichtenberg2022}.
Magma ocean states play a substantial role in establishing the geophysical and climatic regimes of rocky planets~\citep{2020ChEG...80l5594F} because atmospheric volatiles are highly soluble in magmatic liquids~\citep{2021ApJ...909L..22K,Dorn2021}.
Due to these similar formation sequences, it was commonly assumed that the divergence of Venus and Earth -- in particular Venus's water loss -- occurred late in their evolution~\citep{Way2020}.

Yet, \citet{Hamano2013} suggested that the dry conditions on Venus may have been directly inherited from the early magma ocean stage if a strongly infrared-absorbing, condensable species such as water is present on the planet.
For Venus, this is supported by isotopic and petrological evidence from meteoritic samples~\citep{2018SSRv..214...36A,2022Natur.611..245B}.
Once the incident stellar flux exceeds a threshold value, a non-linear transition from mixed condensed and volatile to a pure volatile state occurs, and the resulting massive steam atmosphere limits the outgoing radiation from the (molten) planetary surface~\citep{Ingersoll1969,Kasting1988,2010ppc..book.....P,Goldblatt2013,2015ExA....40..449L,2017JGRE..122.1458S}.
This runaway greenhouse state prevents the planet's rapid cooling and can extend the magma ocean stage to hundreds of \SI{}{\mega\year}~\citep{2016ApJ...829...63S,2021AsBio..21.1325B}, enough to remove the entire water reservoir from a rocky planet by H$_2$O photolysis and subsequent hydrodynamic escape of hydrogen~\citep{2013ApJ...778..154W,2014ApJ...785L..20W,Luger2015}.
Venus' atmospheric composition is consistent with the planet having gone through this phase~\citep{2020NatGe..13..265G}, although the past presence or absence of a Venusian water ocean has not been definitely established~\citep{Raymond2006,Raymond2007,Hamano2013,Way2016,2019JGRE..124.2015K,2021JGRE..12606643K,Turbet2021,2023PNAS..12009751W}.

The runaway greenhouse transition is a robust prediction from climate models~\citep{Forget2014,Boukrouche2021,2022A&A...658A..40C}, and its impact on planetary bulk properties has been shown to be in the detectable range of current astronomical instrumentation~\citep{Goldblatt2015}.
In particular, planets in a super-runaway climate state are expected to be thermally inflated~\citep{Turbet2019,Turbet2020,Mousis2020}, which directly increases their transit radii by an amount that is sensitive on the water content.
However, dissolution of water~\citep[e.g.,][]{Elkins-Tanton2008,Hier-Majumder2017,Salvador2023} in the magma may decrease this effect~\citep{Dorn2021}, and variations in interior chemical exchange between core and mantle material may influence the outgassed volatiles that are visible in the atmosphere via transmission spectroscopy~\citep{2021ApJ...914L...4L,Schlichting2022}.
Astronomical observations of planets that are currently in the super-runaway phase may thus constrain properties of their mantles and establish an observational connection between exoplanetary interiors and atmospheres~\citep{Lichtenberg2022,Wordsworth2022}.

Of particular relevance is that the radiation-induced transition between a runaway greenhouse state and a temperate climate is thought to occur at a relatively sharp threshold instellation~\citep{2010ppc..book.....P,Kopparapu2013}.
Consequently, the instellation at which the runaway greenhouse transition occurs is aptly considered to be the inner boundary of the habitable zone.
Its prevalent definition refers to the possibility of liquid water on the surface of an Earth-like planet with an oxidized CO$_2$/H$_2$O/N$_2$-rich atmosphere~\citep[][]{Kasting1993,Kopparapu2013,Kopparapu2014}, but it can be influenced by different internal redox states and volatile compositions~\citep{2011ApJ...734L..13P,2017ApJ...837L...4R,2018ApJ...858...72R,2019ApJ...875...31K,2020ApJ...896..115G,2022JGRE..12707456G,2023ApJ...942L..20H}.
The fundamental concept of a habitable zone dates back centuries~\citep{Newton1687,Whewell1858,Shapley1953,Huang1959}, and its modern form has proven popular in the planetary literature\footnote{At the time of writing, a search of the term ``habitable zone'' in the titles and abstracts of refereed articles in the National Aeronautics and Space Administration (NASA) Astrophysics Data System returned $\sim\num{1700}$ results.}.
However, it should be emphasized that the habitable zone, as it stands today, is merely a concept based on theoretical predictions and geochemical evidence from one planet -- Earth~\citep{2020SciA....6.1420C} -- and its general validity continues to be debated~\citep[e.g.,][]{Cockell2016,Moore2017,Tuchow2023}.
The question naturally emerges if a planetary habitable zone -- in its common form with boundaries simply defined by the received stellar irradiation -- is a predictive theoretical concept and how the habitable zone hypothesis can be tested observationally.

Observational tests being considered include searches for direct evidence of liquid water conveyed by ocean glint~\citep{Williams2008,Robinson2010,Lustig-Yaeger2018} or water vapor in planetary atmospheres~\citep{Suissa2020}.
A different approach relies on comparative planetology: By aiming for a statistical detection in a planet population, degeneracies and ambiguousities in making statements about the habitability of individual planets may be overcome~\citep{Checlair2019,Apai2019a}.
Tests suggested in the literature include determining, for a range of orbital distances, atmospheric H$_{2}$O and CO$_2$ abundances, planetary albedos~\citep{Bean2017,Bixel2021}, or colors~\citep{Crow2011,Bixel2020}, or testing the relationship between CO$_2$ partial pressure and incident flux~\citep{Lehmer2020}.
All these tests require surveying a large enough sample of terrestrial-sized planets with next-generation instruments, rendering them out of reach in the immediate future.

Here, we explore the feasibility of a statistical test of the habitable zone hypothesis by surveying planetary bulk properties close to its inner edge, the runaway greenhouse transition.
Our goal is to assess the ability of near-future transit surveys to test the hypothesis that the runaway greenhouse effect causes a discontinuity of planetary radii and bulk densities when ordered by receiving instellation~\citep{Turbet2019}.
Our main tool for this is \bioverse, a simulation framework for assessing the statistical power of exoplanet surveys~\citep{Bixel2021}.
It consists of a sample generator that populates stars from the \gaia\ catalog~\citep[][Hardegree-Ullman et al., submitted]{Gaia_mission,Gaia_DR3} with planetary systems based on state-of-the-art occurrence rates~\citep{Bergsten2022}, a flexible survey simulator that allows for a broad range of trade studies, and a hypothesis testing module that quantifies the survey's ability to detect a previously injected trend.
Trying different instrumentation and survey designs, we use \bioverse\ to recover runaway greenhouse-induced effects based on model predictions~\citep{Turbet2020,Dorn2021} that we inject into a baseline planet population.

@@

\section{Baseline Model}\label{sec:met_baseline}
%The purpose of this paper is to demonstrate the feasibility of measuring the demographic feature introduced above with near-future exoplanet missions.
%Central to our procedure is to explore the detectability of population-level trends with future exoplanet surveys, and here the trend in question is the proposed radius discontinuity.
%To enable a quantitative assessment of this detectability, we injected the signal into a simulated planet sample before observing it with simulated surveys.
The goal of this study is to determine -- for different configurations of near-future exoplanet surveys -- the confidence level with which the runaway greenhouse threshold can be detected statistically. %in simulated exoplanet observations.
Our basic methodology was as follows:
We expanded the \bioverse\ framework~\citep{Bixel2020,Bixel2021}\footnote{\bioverse\ is actively maintained and documented open source software written in Python. Its latest version and documentation can be found at \url{https://github.com/danielapai/bioverse}.} to generate synthetic samples of stars that host planets according to the observed exoplanet demographics.
We then adapted planetary bulk properties as predicted from models of runaway greenhouse atmospheres, simulated observations of the planets, and computed Bayesian evidences in favor of a habitable zone inner edge discontinuity (see diagram in Figure~\ref{fig:flowchart}).
\begin{figure*}
    \begin{centering}
        \includegraphics[width=\hsize]{figures/flowchart.pdf}
        \caption{Workflow of our hypothesis testing with \bioverse. First, we generate a sample of stars and populate them with planets based on \kepler\ demographics.
            A fraction of them are then assigned a runaway greenhouse climate based on the model described in Sect.~\ref{sec:met_rghmodel}.
        We then simulate an exoplanet survey, whereby selection effects and detection biases are introduced. Finally, we test the runaway greenhouse hypothesis based on data from the survey simulation.
        By iterating through these steps, we compute the statistical power of testing the hypothesis for different survey designs.}
        \label{fig:flowchart}
    \end{centering}
\end{figure*}
In this section, we review the source of the stellar sample, the modeled luminosity evolution, the generation of a synthetic planet sample, and the orbital parameters of the planets.
An overview of our key assumptions and model parameters can be found in Table~\ref{tab:params_table}.

\subsection{Stellar sample from Gaia DR3}
%Different sample sizes are realized by varying the maximum distance of stars to the solar system $d_\mathrm{max}$.
%\todo[inline]{KEVIN: describe stellar population}
\todo[inline]{KEVIN: update reference to my paper once accepted}
The original \bioverse\ stellar catalog was generated randomly from the \citet{Chabrier2003} stellar mass function. Exquisite parallax and photometric data from the \textit{Gaia} mission made it possible to generate a homogeneous and complete stellar catalog out to about 100~pc, which became the new standard stellar catalog for \bioverse ~(Hardegree-Ullman et al., submitted). Here, we briefly describe how stellar parameters $T_{\mathrm{eff}}$, $L_{\star}$, $R_{\star}$, and $M_{\star}$ were derived.

Hardegree-Ullman et al. (submitted) used the Gaia Catalogue of Nearby Stars~\citep[hereafter GCNS,][]{Smart2021} as the basis for deriving stellar parameters for the \bioverse\ catalog.
The GCNS identified stars out to 120~pc, and includes \textit{Gaia} DR3 parallaxes and photometry in $G$, $G_{BP}$, and $G_{RP}$ bands~\citep{GaiaCollaboration2022} and $K_S$ band photometry from 2MASS~\citep{Cutri2003}, all of which were used in stellar classification.
From this information, colors and absolute magnitudes were computed, and initial color-magnitude cuts were made to remove non-main sequence stars.
Effective temperatures were computed primarily from color-temperature relations derived using the main sequence stellar parameters table from \citet{Pecaut2013}.
Next, luminosities were computed from absolute $G$-band magnitudes and a derived bolometric correction.
Stellar radii were then computed with the effective temperatures and luminosities using the Stefan-Boltzmann law, or using absolute $K_S$-band magnitudes and an empirical radius-luminosity relation from \citet{Mann2015} for targets within the absolute magnitude range of M dwarfs.
Finally, masses were derived from the mass-luminosity relation of \citet{Torres2010} for stars with $M_{\star}\gtrsim 0.7\,M_{\odot}$, and from that of \citet{Mann2019} for targets within the absolute magnitude range of M dwarfs.
The derived stellar parameters from Hardegree-Ullman et al. (submitted) were compared to measured stellar parameters for all known exoplanet hosts from the literature and were found to be consistent within 1\%, 3\%, and 5.5\% for $T_{\mathrm{eff}}$, $R_{\star}$, and $M_{\star}$, respectively, which are all below the typical measurement uncertainties of 3.3\%, 6.8\%, and 7.9\%, respectively.
From this catalog, \bioverse\ samples stars within an isotropic distance from the Solar System as required by the planetary sample size.

\subsection{Stellar luminosity evolution}
Planetary systems are hosted by stars of a wide range of ages, and stellar luminosities evolve with time.
Since the presence of a runaway greenhouse state is highly dependent on the amount of radiation received by the planet, and thus on the luminosity of the host star, we assigned age-dependent luminosities to our synthetic stars.

While stellar ages are notoriously poorly constrained~\cite[e.g.,][]{Adams2005}, the age distribution of planet host stars in the Solar neighborhood was shown to be broadly consistent with uniform~\citep{Reid2007,Gaidos2023}.
For our synthetic stars, we thus drew random ages from a uniform distribution from \SI{0}{\giga\year} to \SI{10}{\giga\year}.
We then assigned each star a luminosity from the mass-dependent evolutionary models in \citet{Baraffe1998}.
Figure~\ref{fig:luminosity_tracks} shows the corresponding luminosity evolution as a function of stellar mass and age.
\begin{figure}[ht!]
    \script{plot_luminosity_tracks.py}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/luminosity_tracks.pdf}
        \caption{
            Bolometric luminosity tracks of stars with different masses, computed from stellar evolution models in \citet{Baraffe1998}.
            Low-mass stars, which make up the majority of stars in the solar neighborhood, undergo an extended early phase of several magnitudes higher luminosity before entering a lifetime of relative faintness.
        }
        \label{fig:luminosity_tracks}
    \end{centering}
\end{figure}

@@

\subsection{Orbit parameters and planet masses}\label{sec:met-orbits_masses}
Eccentric orbits alter the probability of a planet to transit~\citep[e.g.,][]{Barnes2007a}.
The distribution of eccentricities $e$ of exoplanets has been found to resemble a Beta function~\citep{Kipping2013b}, which we chose to draw synthetic eccentricities from.
Following~\citet{Kipping2013b}, we used a Beta distribution with parameters $a=0.867$ and $b=3.03$, and truncated the distribution at $e = 0.8$.
%Another parameter with an even higher impact on transit probabilities is the orbital inclination toward the observer $i$.
Assuming isotropic alignments of orbits, we assigned each planet an inclination drawn from a distribution uniform in $\cos(i)$.

@@

\section{Discussion}\label{sec:discussion}
%\todo[inline]{\textbf{Potential additional discussion points: }\\
%(How close is Earth to runaway? (-> \citet{Goldblatt2012,Popp2016}; what is the max. expected change in Earth's albedo due to deglaciation?)}

>>>>

% Author list
\author[0000-0001-8355-2107]{Martin Schlecker}
\affiliation{Steward Observatory, The University of Arizona, Tucson, AZ 85721, USA; \href{mailto:schlecker@arizona.edu}{schlecker@arizona.edu}}
\author[0000-0003-3714-5855]{D\'{a}niel Apai}
\affiliation{Steward Observatory, The University of Arizona, Tucson, AZ 85721, USA; \href{mailto:schlecker@arizona.edu}{schlecker@arizona.edu}}
\affiliation{Lunar and Planetary Laboratory, The University of Arizona, Tucson, AZ 85721, USA}
\author[0000-0002-3286-7683]{Tim Lichtenberg}
\affiliation{Kapteyn Astronomical Institute, University of Groningen, PO Box 800, 9700 AV Groningen, The Netherlands}
\author[0000-0003-4500-8850]{Galen Bergsten}
\affiliation{Lunar and Planetary Laboratory, The University of Arizona, Tucson, AZ 85721, USA}
\author[0000-0001-8106-6164]{Arnaud Salvador}
\affiliation{Lunar and Planetary Laboratory, The University of Arizona, Tucson, AZ 85721, USA}
\author[0000-0003-3702-0382]{Kevin K.\ Hardegree-Ullman}
\affiliation{Steward Observatory, The University of Arizona, Tucson, AZ 85721, USA; \href{mailto:schlecker@arizona.edu}{schlecker@arizona.edu}}

% Abstract
\begin{abstract}
Long-term magma ocean phases on rocky exoplanets orbiting closer to their star than the runaway greenhouse threshold -- the inner edge of the classical habitable zone -- may offer insights into the physical and chemical processes that distinguish potentially habitable worlds from others.
Thermal stratification of runaway planets is expected to significantly inflate their atmospheres, potentially providing observational access to the runaway greenhouse transition in the form of a ``habitable zone inner edge discontinuity'' in radius--density space.
Here, we use \bioverse, a statistical framework combining contextual information from the overall planet population with a survey simulator, to assess the ability of ground- and space-based telescopes to test this hypothesis.

We find that the demographic imprint of the runaway greenhouse transition is likely detectable with high-precision transit photometry for sample sizes $\gtrsim$~100~planets if at least \SI{\sim 10}{\percent} of those orbiting closer than the habitable zone inner edge harbor runaway climates.
%We find that the demographic imprint of the runaway greenhouse transition is likely detectable with high-precision transit photometry for planet samples of size $\gtrsim$~100 if at least \SI{\sim 10}{\percent} of planets inside of the habitable zone inner edge harbor runaway climates.
Our survey simulations suggest that in the near future, ESA's \plato\ mission will be the most promising survey to probe the habitable zone inner edge discontinuity.
We determine survey strategies that maximize the diagnostic power of the obtained data and identify as key mission design drivers: 1. A follow-up campaign of planetary mass measurements and 2. The fraction of low-mass stars in the target sample.
Observational constraints on the runaway greenhouse transition will provide crucial insights into the distribution of atmospheric volatiles among rocky exoplanets, which may help to identify the nearest potentially habitable worlds.
\end{abstract}

% Main body
\section{Introduction}
Despite recent advancements in observational techniques, our understanding of terrestrial-sized planets remains woefully limited, with fundamental aspects of their nature, composition, and potential habitability still largely unknown.
Due to the inherent biases of current exoplanet detection techniques, the best-studied category of rocky exoplanets at present is that of hot or warm, close-in planets~\citep{2019AREPS..47..141J,2021JGRE..12606639B}.
These experience thermal states that are in some aspects comparable to the ones of the inner solar system bodies at early stages of their evolution~\citep{Ikoma2018,2021ChEG...81l5735C}, which likely profoundly affected the distribution of volatiles between planetary core, mantle, and atmosphere.
Studying the geophysical state of hot exoplanets can thus inform our understanding of the early evolutionary stages of Earth and other habitable worlds~\citep{Lichtenberg2022,Krijt2022}.

An example from the solar system for the potential significance of these early stages are the divergent atmospheric evolutions of Venus and Earth~\citep[e.g.,][]{2019JGRE..124.2015K,2021JGRE..12606643K,Salvador2023b}.
While having accreted from a similar mass reservoir~\citep{2020plas.book..287R,2020SSRv..216...55K,2020SSRv..216...27M,2020plas.book....3Z} and despite their similar bulk properties~\citep{Smrekar2018}, they evolved into planets with very different surface conditions~\citep{1982Sci...216..630D,Kasting1988,Hamano2013,Kane2014,Way2020,Turbet2021}.
Both planets likely underwent a giant impact phase~\citep{2020plas.book..287R,2020NatGe..13..265G,Liu2022} that melted their mantles~\citep{2012AREPS..40..113E,2018RSPTA.37680109S,Lichtenberg2022}.
Magma ocean states play a substantial role in establishing the long-term geophysical and climatic regimes of rocky planets~\citep{2020ChEG...80l5594F}, in particular owing to efficient heat and volatile transfers between interior and atmosphere in the absence of a stiff boundary separating them~\citep{2021ApJ...909L..22K,Dorn2021,Salvador2023b}.
% because atmospheric volatiles are highly soluble in magmatic liquids~\citep{2021ApJ...909L..22K,Dorn2021}.
Due to these similar formation sequences, it was commonly assumed that the divergence of Venus and Earth -- in particular Venus' water loss -- occurred late in their evolution~\citep[e.g.,][]{Way2020}.

Yet, \citet{Hamano2013} suggested that the present-day dry conditions on Venus may have been directly inherited from the early magma ocean stage.% if a strongly infrared-absorbing, condensable species such as water is present on the planet.
If a strongly infrared-absorbing, condensable species such as water was dominant in the atmosphere, the resulting strong thermal blanketing effect would prevent the planet to efficiently radiate to space and maintain the surface molten~\citep{Ingersoll1969,Kasting1988,2010ppc..book.....P,Goldblatt2013,2015ExA....40..449L,2017JGRE..122.1458S}.
This runaway greenhouse state can extend the magma ocean stage to hundreds of \SI{}{\mega\year}~\citep{2016ApJ...829...63S,2021AsBio..21.1325B}, enough to remove the entire water reservoir from a rocky planet by H$_2$O photolysis and subsequent hydrodynamic escape of hydrogen~\citep{2013ApJ...778..154W,2014ApJ...785L..20W,Luger2015}.
%This scenario requires early water delivery, which for Venus is supported by isotopic and petrological evidence from meteoritic samples~\citep{2018SSRv..214...36A,2022Natur.611..245B}.
%Where the incident stellar flux exceeds a threshold value, a non-linear transition from a mixture of condensed and volatile states to a pure volatile state occurs, and the resulting massive steam atmosphere limits the outgoing radiation from the (molten) planetary surface~\citep{Ingersoll1969,Kasting1988,2010ppc..book.....P,Goldblatt2013,2015ExA....40..449L,2017JGRE..122.1458S}.
For Venus, isotopic and petrological evidence from meteoritic samples~\citep{2018SSRv..214...36A,2022Natur.611..245B} and the planet's atmospheric composition suggest that it went through this phase~\citep{2020NatGe..13..265G}.
Although the past presence or absence of a Venusian water ocean has not been definitely established~\citep{Raymond2006,Raymond2007,Hamano2013,Way2016,2019JGRE..124.2015K,2021JGRE..12606643K,Turbet2021,2023PNAS..12009751W}, a transient habitable phase cannot be conclusively ruled out~\citep[e.g.,][]{Way2016,Salvador2017,Krissansen-Totton2021}.

The runaway greenhouse transition is a robust prediction from climate models~\citep{Kasting1988,Nakajima1992,Goldblatt2012,Forget2014,Boukrouche2021,2022A&A...658A..40C}, and its impact on planetary bulk properties has been shown to be in the detectable range of current astronomical instrumentation~\citep{Goldblatt2015}.
In particular, planets in a runaway greenhouse state are expected to be thermally inflated~\citep{Turbet2019,Turbet2020,Mousis2020}, which directly increases their transit radii by an amount that is a function of the water content.
However, dissolution of water~\citep[e.g.,][]{Elkins-Tanton2008,Hier-Majumder2017,Salvador2023} in the magma may decrease this effect~\citep{Dorn2021}, and chemical exchange between core and mantle material may influence the amount and speciation of outgassed volatiles that are visible in the atmosphere via transmission spectroscopy~\citep{2021ApJ...914L...4L,Schlichting2022}.
Astronomical observations of planets that are currently in a runaway greenhouse state may thus constrain properties of their mantles and establish an observational connection between exoplanetary interiors and atmospheres~\citep{Lichtenberg2022,Wordsworth2022}.

Of particular relevance is that the radiation-induced transition between a runaway greenhouse state and a temperate climate is thought to occur at a relatively sharp instellation threshold~\citep{Goldblatt2013,Leconte2013,Kopparapu2013}.
Consequently, the instellation at which the runaway greenhouse transition occurs is aptly considered to be the inner boundary of the habitable zone~\citep[e.g.,][]{Ramirez2018,Salvador2023b}.
Its prevalent definition refers to the possibility of a sustained liquid water body on the surface of an Earth-like planet with an oxidized CO$_2$/H$_2$O/N$_2$-rich atmosphere~\citep[][]{Kasting1993,Kopparapu2013,Kopparapu2014};
its exact spatial location and extent may be strongly influenced by the interior and atmosphere oxidation state and resulting atmosphere composition~\citep{2011ApJ...734L..13P,2017ApJ...837L...4R,2018ApJ...858...72R,2019ApJ...875...31K,2020ApJ...896..115G,2022JGRE..12707456G,2023ApJ...942L..20H}.
The fundamental concept of a habitable zone dates back centuries~\citep{Newton1687,Whewell1858,Shapley1953,Huang1959}, and its modern form has proven popular in the planetary literature\footnote{At the time of writing, a search of the term ``habitable zone'' in the titles and abstracts of refereed articles in the National Aeronautics and Space Administration (NASA) Astrophysics Data System returned $\sim\num{1700}$ results.}.
However, it should be emphasized that the habitable zone, as it stands today, is merely a concept based on theoretical predictions and geochemical evidence from one planet -- Earth~\citep{2020SciA....6.1420C} -- and its general validity remains controversial~\citep[e.g.,][]{Cockell2016,Moore2017,Tuchow2023}.
The question naturally emerges if a planetary habitable zone -- in its common form with boundaries defined by stellar irradiation -- is a predictive theoretical concept and how the habitable zone hypothesis can be tested observationally.

Observational tests being considered include searches for direct evidence of liquid water conveyed by ocean glint~\citep{Williams2008,Robinson2010,Lustig-Yaeger2018} or water vapor in planetary atmospheres~\citep{Suissa2020}.
A different approach relies on comparative planetology:
aiming for a statistical detection in a planet population provides robustness against ambiguity that could otherwise arise from individual variations in a planet's composition or geophysical history~\citep{Checlair2019,Apai2019a}.
Tests suggested in the literature include determining, for a range of orbital distances, atmospheric H$_{2}$O and CO$_2$ abundances, planetary albedos~\citep{Bean2017,Bixel2021}, or colors~\citep{Crow2011,Bixel2020}, or testing the relationship between CO$_2$ partial pressure and incident flux~\citep{Lehmer2020}.
All these tests require surveying a large enough sample of terrestrial-sized planets with next-generation instruments, rendering them out of reach in the immediate future.

Here, we explore the feasibility of a statistical test of the habitable zone hypothesis by surveying planetary bulk properties close to its inner edge, the runaway greenhouse transition.
Our goal is to assess the ability of near-future transit surveys to test the hypothesis that the runaway greenhouse effect causes a discontinuity of planetary radii and bulk densities when ordered by receiving instellation~\citep{Turbet2019}.
Our main tool for this is \bioverse, a simulation framework for assessing the statistical power of exoplanet surveys~\citep{Bixel2021}.
It consists of a sample generator that populates stars from the \gaia\ catalog~\citep[][]{Hardegree-Ullman2023,Gaia_mission,Gaia_DR3} with planetary systems based on state-of-the-art occurrence rates~\citep{Bergsten2022}, a flexible survey simulator that allows for a broad range of trade studies, and a hypothesis testing module that quantifies the survey's ability to detect a previously injected trend.
Trying different instrumentation and survey designs, we use \bioverse\ to recover runaway greenhouse-induced effects based on model predictions~\citep{Turbet2020,Dorn2021} that we inject into a baseline planet population.

@@

\section{Baseline Model}\label{sec:met_baseline}
The goal of this study is to determine -- for different configurations of near-future exoplanet surveys -- the confidence level with which the runaway greenhouse threshold can be detected statistically. %in simulated exoplanet observations.
Our basic methodology was as follows:
We expanded the \bioverse\ framework~\citep{Bixel2020,Bixel2021}\footnote{\bioverse\ is actively maintained and documented open source software written in Python. Its latest version and documentation can be found at \url{https://github.com/danielapai/bioverse}.} to generate synthetic samples of stars that host planets according to the observed exoplanet demographics.
We then adapted planetary bulk properties as predicted from models of runaway greenhouse atmospheres, simulated observations of the planets, and computed Bayesian evidences in favor of a habitable zone inner edge discontinuity (see diagram in Figure~\ref{fig:flowchart}).
\begin{figure*}
    \begin{centering}
        \includegraphics[width=\hsize]{figures/flowchart.pdf}
        \caption{Workflow of our hypothesis testing with \bioverse. First, we generate a sample of stars and populate them with planets based on \kepler\ demographics.
            A fraction of them are then assigned a runaway greenhouse climate based on the model described in Sect.~\ref{sec:met_rghmodel}.
        We then simulate an exoplanet survey, whereby selection effects and detection biases are introduced. Finally, we test the runaway greenhouse hypothesis based on data from the survey simulation.
        By iterating through these steps, we compute the statistical power of testing the hypothesis for different survey designs.}
        \label{fig:flowchart}
    \end{centering}
\end{figure*}
In this section, we review the source of the stellar sample, the modeled luminosity evolution, the generation of a synthetic planet sample, and the orbital parameters of the planets.
An overview of our key assumptions and model parameters can be found in Table~\ref{tab:params_table}.

\subsection{Stellar sample from \gaia\ DR3}
The original \bioverse\ stellar catalog was generated randomly from the \citet{Chabrier2003} stellar mass function.
Improved parallax and photometric data from the \gaia\ mission made it possible to generate a homogeneous and complete stellar catalog out to about 100~pc, which became the new standard stellar catalog for \bioverse ~\citep{Hardegree-Ullman2023}.
Here, we briefly describe how we derived the stellar effective temperature $T_{\mathrm{eff}}$, luminosity $L_{\star}$, stellar radius $R_{\star}$, and stellar mass $M_{\star}$.

\citet{Hardegree-Ullman2023} used the \gaia\ Catalogue of Nearby Stars~\citep[hereafter GCNS,][]{Smart2021} as the basis for deriving stellar parameters for the \bioverse\ catalog.
The GCNS identified stars out to 120~pc and includes \gaia\ DR3 parallaxes and photometry in $G$, $G_{BP}$, and $G_{RP}$ bands~\citep{GaiaCollaboration2022} and $K_S$ band photometry from 2MASS~\citep{Cutri2003}, all of which were used in stellar classification.
From this information, we computed colors and absolute magnitudes and applied initial color-magnitude cuts to remove non-main sequence stars.
We derived effective temperatures primarily from color-temperature relations derived using the main sequence stellar parameters table from \citet{Pecaut2013}.
Luminosities were computed from absolute $G$-band magnitudes and a derived bolometric correction.
We computed stellar radii with the effective temperatures and luminosities using the Stefan-Boltzmann law or using absolute $K_S$-band magnitudes and an empirical radius-luminosity relation from \citet{Mann2015} for targets within the absolute magnitude range of M~dwarfs.
Finally, we derived masses from the mass-luminosity relation of \citet{Torres2010} for stars with $M_{\star}\gtrsim 0.7\,M_{\odot}$, and from that of \citet{Mann2019} for targets within the absolute magnitude range of M dwarfs.
The derived stellar parameters from \citet{Hardegree-Ullman2023} were compared to measured stellar parameters for all known exoplanet hosts from the literature and were found to be consistent within 1\%, 3\%, and 5.5\% for $T_{\mathrm{eff}}$, $R_{\star}$, and $M_{\star}$, respectively, which are all below the typical measurement uncertainties of 3.3\%, 6.8\%, and 7.9\%, respectively.
From this catalog, \bioverse\ samples stars within an isotropic distance from the solar system as required by the planetary sample size.

\subsection{Stellar luminosity evolution}
Planetary systems are hosted by stars of a wide range of ages, and stellar luminosities evolve with time.
Since the occurrence of a runaway greenhouse state is highly dependent on the amount of radiation received by the planet, and thus on the luminosity of the host star, we assigned age-dependent luminosities to our synthetic stars.

While stellar ages are notoriously poorly constrained~\cite[e.g.,][]{Adams2005}, the age distribution of planet host stars in the Solar neighborhood was shown to be broadly consistent with uniform~\citep{Reid2007,Gaidos2023}.
For our synthetic stars, we thus drew random ages from a uniform distribution from \SI{0}{\giga\year} to \SI{10}{\giga\year}.
We then assigned each star a luminosity from the mass-dependent evolutionary models of \citet{Baraffe1998}.
Figure~\ref{fig:luminosity_tracks} shows the corresponding luminosity evolution as a function of stellar mass and age.
\begin{figure}[ht!]
    \script{plot_luminosity_tracks.py}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/luminosity_tracks.pdf}
        \caption{
            Bolometric luminosity tracks of stars with different masses, computed from stellar evolution models of \citet{Baraffe1998}.
            Low-mass stars, which make up the majority of stars in the solar neighborhood, undergo an extended early phase of several magnitudes higher luminosity before entering a lifetime of relative faintness.
        }
        \label{fig:luminosity_tracks}
    \end{centering}
\end{figure}

@@

\subsection{Orbit parameters and planet masses}\label{sec:met-orbits_masses}
Eccentric orbits alter the probability of a planet to transit~\citep[e.g.,][]{Barnes2007a}.
The distribution of eccentricities $e$ of exoplanets has been found to resemble a Beta function~\citep{Kipping2013b}, which we chose to draw synthetic eccentricities from.
Following~\citet{Kipping2013b}, we used a Beta distribution with parameters $a=0.867$ and $b=3.03$, and truncated the distribution at $e = 0.8$.
Assuming isotropic alignments of orbits, we assigned each planet an inclination drawn from a distribution uniform in $\cos(i)$.

@@

\section{Discussion}\label{sec:discussion}
%(How close is Earth to runaway? (-> \citet{Goldblatt2012,Popp2016}; what is the max. expected change in Earth's albedo due to deglaciation?)}
